### 2024-09-24 paper/paper.qmd
intital draft update

---
title: "Traffic Collisions in Toronto: Analysis of Key Variables"
subtitle: "An Analysis of Traffic Collision Data from 2014-2024"
author: 
  - Andy Jiang
thanks: "Code and data are available at: https://github.com/AndyYanxunJiang/toronto-traffic-collisions.git"
date: 24 September 2024
date-format: long
abstract: "This paper analyzes traffic collision data in Toronto from 2014 to 2024. It explores trends in collisions by month, year, and neighborhood, examining variables such as fatalities, injuries, and property damage. The analysis is done using R, with visualizations produced through ggplot2."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)

# Load the cleaned data
data <- read_csv(here::here("data/analysis_data/cleaned_collision_data.csv"))

# Optional: Convert relevant variables to factors
data <- data %>%
  mutate(OCC_MONTH = factor(OCC_MONTH, levels = month.name),
         INJURY_COLLISIONS = as.factor(INJURY_COLLISIONS),
         FTR_COLLISIONS = as.factor(FTR_COLLISIONS),
         PD_COLLISIONS = as.factor(PD_COLLISIONS),
         AUTOMOBILE = as.factor(AUTOMOBILE),
         PEDESTRIAN = as.factor(PEDESTRIAN))
```

# Introduction
This paper analyzes traffic collision data in Toronto from 2014 to 2024. We explore trends in monthly, yearly, and neighborhood-level collisions, with a focus on variables such as fatalities, injuries, property damage, and modes of transport involved. The data is sourced from Open Data Toronto and contains detailed information about each collision.

The remainder of this paper is structured as follows. Section @sec-data provides details of the dataset used in this analysis. Section @sec-results presents the results of the analysis, followed by a discussion in Section @sec-discussion. Finally, Section @sec-conclusion summarizes the key findings of the study.

# Data {#sec-data}
The dataset used in this analysis was obtained from Open Data Toronto. It includes traffic collision data collected between 2014 and 2024. The variables include the date and time of the collision, whether there were fatalities, injuries, or property damage, and the type of vehicle(s) involved. @tbl-1 provides a sample of the cleaned dataset.

```{r}
#| label: tbl-1
#| tbl-cap: Sample of the cleaned collision data
#| echo: false

head(data, 10) |>
  kable(
    col.names = c("Month", "Day of Week", "Year", "Hour", "Fatalities",
                  "Injury Collisions", "FTR Collisions", "PD Collisions", 
                  "Neighbourhood", "Automobile", "Motorcycle", 
                  "Passenger", "Bicycle", "Pedestrian"),
    booktabs = TRUE
  )
```

# Results {#sec-results}
Here we analyze the data by looking at the frequency of traffic collisions based on various factors such as month, year, and neighborhood.

Collisions by Month
The following figure shows the distribution of collisions by month across all years.

```{r}
#| label: fig-collisions-month
#| fig-cap: Collisions by Month
#| echo: false

data %>% 
  ggplot(aes(x = OCC_MONTH, fill = factor(OCC_YEAR))) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  labs(title = "Traffic Collisions by Month (2014-2024)",
       x = "Month", y = "Count", fill = "Year") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Collisions by Hour of the Day

```{r}
#| label: fig-collisions-hour
#| fig-cap: Collisions by Hour of the Day
#| echo: false

data %>% 
  ggplot(aes(x = OCC_HOUR, fill = factor(OCC_YEAR))) +
  geom_histogram(binwidth = 1, position = "dodge") +
  theme_minimal() +
  labs(title = "Traffic Collisions by Hour of the Day (2014-2024)",
       x = "Hour", y = "Count", fill = "Year")

```

```{r}
#| label: fig-monthly-collisions
#| fig-cap: Monthly distribution of traffic collisions
#| echo: false

data %>%
  count(OCC_MONTH) %>%
  ggplot(aes(x = OCC_MONTH, y = n)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme_minimal() +
  labs(title = "Monthly Distribution of Traffic Collisions", 
       x = "Month", y = "Number of Collisions")

```

```{r}
#| label: fig-fatalities-by-year
#| fig-cap: Yearly trend in fatalities from collisions
#| echo: false

data %>%
  group_by(OCC_YEAR) %>%
  summarize(Total_Fatalities = sum(FATALITIES, na.rm = TRUE)) %>%
  ggplot(aes(x = OCC_YEAR, y = Total_Fatalities)) +
  geom_line(color = "darkred", size = 1) +
  theme_minimal() +
  labs(title = "Fatalities by Year", x = "Year", y = "Number of Fatalities")

```
\newpage
# Discussion {#sec-discussion}
Overall, the analysis of traffic collisions from 2014 to 2024 reveals several key trends. Traffic collisions are more frequent during certain months of the year, with higher occurrences during the winter months, potentially due to hazardous road conditions. The distribution of collisions by hour of the day indicates that rush hours experience the highest number of collisions. The relationship between fatalities and injury collisions suggests that even non-fatal collisions result in significant injuries, emphasizing the need for road safety interventions.

# Conclusion {#sec-conclusion}
This analysis highlights the importance of studying traffic collision data to inform policy decisions and road safety measures. Further analysis could include a more detailed breakdown by specific neighborhoods or examining the impact of weather conditions on collision frequency.

\newpage

# References

### 2024-09-24 paper/temp.qmd
intital draft update

--- 2024-09-24 paper/paper.qmd
@@
 thanks: "Code and data are available at: https://github.com/AndyYanxunJiang/toronto-traffic-collisions.git"
-date: 24 September 2024
+date: today
 date-format: long

### 2024-09-24 scripts/02-data_cleaning.R

@@ -21,3 +21,4 @@ cleaned_data <- raw_data %>%
 
 #### Save the cleaned data ####
 write_csv(cleaned_data, "data/analysis_data/cleaned_collision_data.csv")
+

### 2024-09-26 scripts/00-simulate_data.R
re-run entire script, update preamble and code. simulated new data, cleaned out more unneeded columns

@@ -1,10 +1,9 @@
 #### Preamble ####
-# Purpose: Simulates traffic collision data for analysis.
+# Purpose: Simulates traffic collision data for pre-analysis
 # Author: Andy Jiang
 # Date: 22 September 2024
 # Contact: ayx.jiang@mail.utoronto.ca
 # Pre-requisites: None
-# Any other information needed? None
 
 #### Workspace setup ####
 library(tidyverse)
@@ -13,13 +12,13 @@ library(tidyverse)
 set.seed(304)
 
 # Define the start and end date
-start_date <- as.Date("2018-01-01")
-end_date <- as.Date("2023-12-31")
+start_date <- as.Date("2014-01-01")
+end_date <- as.Date("2024-12-31")
 
-# Set the number of random dates you want to generate
+# Set the number of random dates
 number_of_dates <- 100
 
-# Simulate a dataset of 100 random traffic collision observations.
+# Simulate 100 random traffic collision observations.
 
 # Data features:
 # - 'division' indicating the police division where the collision occurred
@@ -28,21 +27,74 @@ number_of_dates <- 100
 # NOTE: Fail to Remain (FTR) collisions and property damage (PD) collisions
 # - Binary indicators for 'automobile', 'pedestrian', and 'bicycle'
 #   involvement in the collisions
+
 simulated_data <- tibble(
   dates = as.Date(
     runif(n = number_of_dates,
           min = as.numeric(start_date), max = as.numeric(end_date)),
     origin = "1970-01-01"
   ),
+  # Assign hour between 0 and 23
+  hour = sample(0:23, number_of_dates, replace = TRUE),
+  
+  # Assign random police divisions
   division = sample(1:14, number_of_dates, replace = TRUE),
-  fatalities = rpois(n = number_of_dates, lambda = 0.05),
-  injury_collisions = rpois(n = number_of_dates, lambda = 1),
-  ftr_collisions = rpois(n = number_of_dates, lambda = 0.5),
-  pd_collisions = rpois(n = number_of_dates, lambda = 2),
-  automobile = sample(c(0, 1), number_of_dates, replace = TRUE),
-  pedestrian = sample(c(0, 1), number_of_dates, replace = TRUE),
-  bicycle = sample(c(0, 1), number_of_dates, replace = TRUE)
+  
+  # Generate fatalities
+  fatalities = {
[…]

### 2024-09-26 scripts/01-download_data.R

@@ -4,7 +4,6 @@
 # Date: 22 September 2024
 # Contact: ayx.jiang@mail.utoronto.ca
 # Pre-requisites: None
-# Any other information needed? None
 
 #### Workspace setup ####
 library(opendatatoronto)

### 2024-09-26 scripts/02-data_cleaning.R

@@ -1,10 +1,9 @@
 #### Preamble ####
-# Purpose: Clean the traffic collision data
+# Purpose: Clean the traffic collision raw data
 # Author: Andy Jiang
 # Date: 23 September 2024
 # Contact: ayx.jiang@mail.utoronto.ca
-# Pre-requisites: Raw data downloaed
-# Any other information needed? None
+# Pre-requisites: Raw data downloaded
 
 #### Workspace setup ####
 library(tidyverse)
@@ -14,11 +13,13 @@ raw_data <- read_csv("data/raw_data/raw_collision_data.csv")
 
 #### Clean data by removing specified columns ####
 # Specific date(day) and precise location removed, not needed for my analysis.
-# id, police division and neighbourhood number removed, useless info.
+# id, police division and neighbourhood removed, analysis not involving
+# location specific relationships.
+# Vehicle type removed due to dominantly being automobile
 cleaned_data <- raw_data %>%
   select(-...1, -OCC_DATE, -X_id, -DIVISION, -HOOD_158, -LONG_WGS84,
-         -LAT_WGS84, -geometry)
+         -LAT_WGS84, -geometry, -NEIGHBOURHOOD_158, -AUTOMOBILE, -MOTORCYCLE,
+         -PASSENGER, -BICYCLE, -PEDESTRIAN)
 
 #### Save the cleaned data ####
 write_csv(cleaned_data, "data/analysis_data/cleaned_collision_data.csv")
-

### 2024-09-26 scripts/03-test_data.R

@@ -3,8 +3,7 @@
 # Author: Andy Jiang
 # Date: 23 September 2024
 # Contact: ayx.jiang@mail.utoronto.ca
-# Pre-requisites: Need to have cleaned data
-# Any other information needed? None
+# Pre-requisites: Need to have cleaned the data
 
 #### Workspace setup ####
 library(tidyverse)
@@ -12,18 +11,27 @@ library(tidyverse)
 #### Load cleaned data ####
 data <- read_csv("data/analysis_data/cleaned_collision_data.csv")
 
-#### Test for negative number of collisions ####
-negative_check <- all(data$FATALITIES >= 0, na.rm = TRUE) &
-  all(data$INJURY_COLLISIONS >= 0, na.rm = TRUE) &
-  all(data$FTR_COLLISIONS >= 0, na.rm = TRUE) &
-  all(data$PD_COLLISIONS >= 0, na.rm = TRUE)
-
 ### Test if fatalities are either NA or positive integers ###
 valid_fatalities_check <- all(is.na(data$FATALITIES) | data$FATALITIES >= 0)
 
-### Test for valid neighborhood names ###
-neighborhood_check <- all(!is.na(data$NEIGHBOURHOOD_158) | 
-                            data$NEIGHBOURHOOD_158 == "NSA")
-
 ### Test for valid collision hour (0 to 23) ###
 valid_hour_check <- all(data$OCC_HOUR >= 0 & data$OCC_HOUR <= 23, na.rm = TRUE)
+
+### Test if `INJURY_COLLISIONS`, `FTR_COLLISIONS`, and `PD_COLLISIONS` 
+### contain only "YES", "NO", or "None" ###
+valid_collision_check <- all(
+  data$INJURY_COLLISIONS %in% c("YES", "NO", "None") &
+    data$FTR_COLLISIONS %in% c("YES", "NO", "None") &
+    data$PD_COLLISIONS %in% c("YES", "NO", "None")
+)
+
+### Test if `OCC_YEAR` is from 2014 to 2024 ###
+valid_year_check <- all(data$OCC_YEAR >= 2014 & 
+                          data$OCC_YEAR <= 2024, na.rm = TRUE)
+
+### Test if `OCC_MONTH` contains valid month names ###
+valid_month_names <- c("January", "February", "March", "April", "May", "June", 
+                       "July", "August", "September", "October", "November", 
+                       "December")
+
+valid_month_check <- all(data$OCC_MONTH %in% valid_month_names, na.rm = TRUE)

### 2024-09-26 paper/paper.qmd
update paper

@@
 It explores trends in collisions by month, year, and neighborhood, examining variables such as fatalities, injuries, and property damage.
-The analysis is done using R, with visualizations produced through ggplot2."
+The analysis is done using @citeR, with visualizations produced through ggplot2."
 format: pdf
@@
 library(knitr)
+library(kableExtra)
+library(opendatatoronto)
+library(here)
+library(tidyr)
+library(dplyr)
+library(janitor)
+library(VennDiagram)
 
@@
 
-# Optional: Convert relevant variables to factors
+# Convert relevant variables to factors
 data <- data %>%
@@
          FTR_COLLISIONS = as.factor(FTR_COLLISIONS),
-         PD_COLLISIONS = as.factor(PD_COLLISIONS),
-         AUTOMOBILE = as.factor(AUTOMOBILE),
-         PEDESTRIAN = as.factor(PEDESTRIAN))
+         PD_COLLISIONS = as.factor(PD_COLLISIONS))
 ```
@@
 
-head(data, 10) |>
+# Display the cleaned collision data
+head(data, 10) %>%
+
+  # Use kable to format the table with custom column names
   kable(
-    col.names = c("Month", "Day of Week", "Year", "Hour", "Fatalities",
-                  "Injury Collisions", "FTR Collisions", "PD Collisions", 
-                  "Neighbourhood", "Automobile", "Motorcycle", 
-                  "Passenger", "Bicycle", "Pedestrian"),
-    booktabs = TRUE
+    col.names = c(
+      "Month", "Day of Week", "Year", "Hour", "Fatalities",
+      "Injury Collisions", "FTR Collisions", "PD Collisions"
+    ),
+    align = 'c'
   )
@@
 #| label: fig-collisions-month
-#| fig-cap: Collisions by Month
+#| fig-cap: Monthly traffic collision counts in Toronto from 2014 to 2024.
 #| echo: false
-
-data %>% 
-  ggplot(aes(x = OCC_MONTH, fill = factor(OCC_YEAR))) +
-  geom_bar(position = "dodge") +
+ggplot(data, aes(x = OCC_MONTH)) +
+  geom_bar(position = "dodge", stat = "count", fill = "lightblue") +
   theme_minimal() +
-  labs(title = "Traffic Collisions by Month (2014-2024)",
-       x = "Month", y = "Count", fill = "Year") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  labs(title = "Total Traffic Collisions by Month (2014-2024)",
+       x = "Month", y = "Total Count of Collisions") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black"),
+        plot.title = element_text(hjust = 0.5, size = 14))
 ```
-
-Collisions by Hour of the Day
 
@@
 #| label: fig-collisions-hour
-#| fig-cap: Collisions by Hour of the Day
+#| fig-cap: Total collisions by hour of the day from 2014 to 2024
 #| echo: false
 
+data$OCC_HOUR[is.na(data$OCC_HOUR)] <- 0
 data %>%
-  ggplot(aes(x = OCC_HOUR, fill = factor(OCC_YEAR))) +
-  geom_histogram(binwidth = 1, position = "dodge") +
+  group_by(OCC_HOUR) %>%
+  summarise(Total_Collisions = sum(INJURY_COLLISIONS == "YES" | 
+                                     FTR_COLLISIONS == "YES" | 
+                                     PD_COLLISIONS == "YES", na.rm = TRUE)) %>%
+  ggplot(aes(x = OCC_HOUR, y = Total_Collisions)) +
+  geom_line(color = "steelblue", size = 1) +
   theme_minimal() +
-  labs(title = "Traffic Collisions by Hour of the Day (2014-2024)",
-       x = "Hour", y = "Count", fill = "Year")
+  scale_x_continuous(breaks = 0:23) +
+  labs(title = "Total Traffic Collisions by Hour of the Day (2014-2024)",
+       x = "Hour of the Day",
+       y = "Total Number of Collisions") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1),
+        plot.title = element_text(hjust = 0.5),
+        panel.grid.major.x = element_line(size = 1.2),
+        panel.grid.major.y = element_line(size = 0.5))
+```
+
+```{r}
 
@@
 ```{r}
-#| label: fig-monthly-collisions
-#| fig-cap: Monthly distribution of traffic collisions
+#| label: fig-collisions-type
+#| fig-cap: Proportion of each type of traffic collisions (2014–2024).
 #| echo: false
 
-data %>%
-  count(OCC_MONTH) %>%
-  ggplot(aes(x = OCC_MONTH, y = n)) +
-  geom_bar(stat = "identity", fill = "skyblue") +
-  theme_minimal() +
-  labs(title = "Monthly Distribution of Traffic Collisions", 
-       x = "Month", y = "Number of Collisions")
+collision_summary <- data %>%
+  summarise(Injury_Collisions = sum(INJURY_COLLISIONS == "YES"),
+            FTR_Collisions = sum(FTR_COLLISIONS == "YES"),
+            PD_Collisions = sum(PD_COLLISIONS == "YES"))
 
+collision_summary_long <- collision_summary %>%
+  pivot_longer(cols = everything(), names_to = "Collision_Type", 
+               values_to = "Count") %>%
+  mutate(Collision_Type = recode(Collision_Type, 
+                                 "Injury_Collisions" = "Injury Collisions",
+                                 "FTR_Collisions" = 
+                                   "Failed to Remain Collisions",
+                                 "PD_Collisions" = 
+                                   "Property Damage Collisions"))
+
+collision_summary_long <- collision_summary_long %>%
+  mutate(Percentage = round(Count / sum(Count) * 100, 2))
+
+ggplot(collision_summary_long, aes(x = "", y = Count, fill = Collision_Type)) +
+  geom_bar(stat = "identity", width = 1) +
+  coord_polar("y", start = 0) +
+  geom_text(aes(label = paste0(Percentage, "%")), 
+            position = position_stack(vjust = 0.5)) +
+  scale_fill_manual(values = c("Injury Collisions" = "lightblue", 
+                               "Failed to Remain Collisions" = "skyblue", 
+                               "Property Damage Collisions" = "#4682B4")) +
+  theme_void() +
+  labs(title = "Proportion of Each Collision Type") +
+  theme(legend.title = element_blank(),
+        plot.title = element_text(hjust = 0.5))
 ```
@@
   geom_line(color = "darkred", size = 1) +
+  scale_x_continuous(breaks = seq(min(data$OCC_YEAR), 
+                                  max(data$OCC_YEAR), by = 1)) +  
   theme_minimal() +
-  labs(title = "Fatalities by Year", x = "Year", y = "Number of Fatalities")
+  labs(title = "Fatalities by Year", 
+       x = "Year", 
+       y = "Number of Fatalities") +
+  theme(
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    plot.title = element_text(hjust = 0.5, size = 14),
+    panel.grid.major.x = element_line(size = 0.8, color = "lightgrey")
+  )
+```
+
+```{r}
+
+```
+
+```{r}
+
+```
+
+```{r}
 

### 2024-09-27 scripts/00-simulate_data.R
styled using styler

@@ -30,16 +30,18 @@ number_of_dates <- 100
 
 simulated_data <- tibble(
   dates = as.Date(
-    runif(n = number_of_dates,
-          min = as.numeric(start_date), max = as.numeric(end_date)),
+    runif(
+      n = number_of_dates,
+      min = as.numeric(start_date), max = as.numeric(end_date)
+    ),
     origin = "1970-01-01"
   ),
   # Assign hour between 0 and 23
   hour = sample(0:23, number_of_dates, replace = TRUE),
-  
+
   # Assign random police divisions
   division = sample(1:14, number_of_dates, replace = TRUE),
-  
+
   # Generate fatalities
   fatalities = {
     temp <- rpois(n = number_of_dates, lambda = 0.05)
@@ -47,20 +49,26 @@ simulated_data <- tibble(
   },
 
   # Generate collisions
-  injury_collisions = ifelse(rpois(n = number_of_dates, lambda = 1) > 0, 
-                             "YES", "NO"),
-  ftr_collisions = ifelse(rpois(n = number_of_dates, lambda = 0.5) > 0, 
-                          "YES", "NO"),
-  pd_collisions = ifelse(rpois(n = number_of_dates, lambda = 2) > 0, 
-                         "YES", "NO"),
-  
+  injury_collisions = ifelse(rpois(n = number_of_dates, lambda = 1) > 0,
+    "YES", "NO"
+  ),
+  ftr_collisions = ifelse(rpois(n = number_of_dates, lambda = 0.5) > 0,
+    "YES", "NO"
+  ),
+  pd_collisions = ifelse(rpois(n = number_of_dates, lambda = 2) > 0,
+    "YES", "NO"
+  ),
+
   # Generate involvement of vehicle or person
-  automobile = ifelse(sample(c(0, 1), number_of_dates, replace = TRUE) == 1, 
-                      "YES", "NO"),
-  pedestrian = ifelse(sample(c(0, 1), number_of_dates, replace = TRUE) == 1, 
-                      "YES", "NO"),
-  bicycle = ifelse(sample(c(0, 1), number_of_dates, replace = TRUE) == 1, 
-                   "YES", "NO")
+  automobile = ifelse(sample(c(0, 1), number_of_dates, replace = TRUE) == 1,
+    "YES", "NO"
+  ),
+  pedestrian = ifelse(sample(c(0, 1), number_of_dates, replace = TRUE) == 1,
+    "YES", "NO"
+  ),
[…]

### 2024-09-27 scripts/01-download_data.R

@@ -18,10 +18,10 @@ resources <- list_package_resources("ec53f7b2-769b-4914-91fe-a37ee27a90b3")
 
 # identify datastore resources; by default, Toronto Open Data sets datastore
 # resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c('csv'))
+datastore_resources <- filter(resources, tolower(format) %in% c("csv"))
 
 # load the first datastore resource as a sample
-collision_data <- filter(datastore_resources, row_number()==1) %>%
+collision_data <- filter(datastore_resources, row_number() == 1) %>%
   get_resource()
 
 # Save the data

### 2024-09-27 scripts/02-data_cleaning.R

@@ -17,9 +17,11 @@ raw_data <- read_csv("data/raw_data/raw_collision_data.csv")
 # location specific relationships.
 # Vehicle type removed due to dominantly being automobile
 cleaned_data <- raw_data %>%
-  select(-...1, -OCC_DATE, -X_id, -DIVISION, -HOOD_158, -LONG_WGS84,
-         -LAT_WGS84, -geometry, -NEIGHBOURHOOD_158, -AUTOMOBILE, -MOTORCYCLE,
-         -PASSENGER, -BICYCLE, -PEDESTRIAN)
+  select(
+    -...1, -OCC_DATE, -X_id, -DIVISION, -HOOD_158, -LONG_WGS84,
+    -LAT_WGS84, -geometry, -NEIGHBOURHOOD_158, -AUTOMOBILE, -MOTORCYCLE,
+    -PASSENGER, -BICYCLE, -PEDESTRIAN
+  )
 
 #### Save the cleaned data ####
 write_csv(cleaned_data, "data/analysis_data/cleaned_collision_data.csv")

### 2024-09-27 scripts/03-test_data.R

@@ -17,7 +17,7 @@ valid_fatalities_check <- all(is.na(data$FATALITIES) | data$FATALITIES >= 0)
 ### Test for valid collision hour (0 to 23) ###
 valid_hour_check <- all(data$OCC_HOUR >= 0 & data$OCC_HOUR <= 23, na.rm = TRUE)
 
-### Test if `INJURY_COLLISIONS`, `FTR_COLLISIONS`, and `PD_COLLISIONS` 
+### Test if `INJURY_COLLISIONS`, `FTR_COLLISIONS`, and `PD_COLLISIONS`
 ### contain only "YES", "NO", or "None" ###
 valid_collision_check <- all(
   data$INJURY_COLLISIONS %in% c("YES", "NO", "None") &
@@ -30,8 +30,10 @@ valid_year_check <- all(data$OCC_YEAR >= 2014 &
                           data$OCC_YEAR <= 2024, na.rm = TRUE)
 
 ### Test if `OCC_MONTH` contains valid month names ###
-valid_month_names <- c("January", "February", "March", "April", "May", "June", 
-                       "July", "August", "September", "October", "November", 
-                       "December")
+valid_month_names <- c(
+  "January", "February", "March", "April", "May", "June",
+  "July", "August", "September", "October", "November",
+  "December"
+)
 
 valid_month_check <- all(data$OCC_MONTH %in% valid_month_names, na.rm = TRUE)

### 2024-09-27 paper/paper.qmd
update paper, styled using styler

--- 2024-09-24 paper/paper.qmd
@@
 thanks: "Code and data are available at: https://github.com/AndyYanxunJiang/toronto-traffic-collisions.git"
-date: 24 September 2024
+date: 26 September 2024
 date-format: long
-abstract: "This paper analyzes traffic collision data in Toronto from 2014 to 2024.
-It explores trends in collisions by month, year, and neighborhood, examining variables such as fatalities, injuries, and property damage.
-The analysis is done using R, with visualizations produced through ggplot2."
+abstract: "This paper analyzes traffic collisions in Toronto from 2014 to 2024, focusing on patterns across months, days of the week, times of day, and types of collisions, including injuries, property damage, and failed to remain incidents.
+Fatalities are analyzed separately as a distinct trend.
+The data, sourced from Open Data Toronto, reveals a light W-shaped trend in monthly collisions, with peaks in January and June.
+Collisions are most frequent between 3 PM and 5 PM, with Friday being the highest day for collisions and Sunday the lowest.
+Property damage accounts for about 71% of all incidents.
+The number of fatalities has generally declined over the years, though trends fluctuate annually.
+These findings highlight the need for ongoing road safety initiatives and targeted interventions during high-risk periods to improve long-term traffic safety."
 format: pdf
-number-sections: true
 bibliography: references.bib
@@
 library(knitr)
+library(opendatatoronto)
+library(dplyr)
 
@@
 
-# Optional: Convert relevant variables to factors
+# Convert relevant variables to factors
 data <- data %>%
-  mutate(OCC_MONTH = factor(OCC_MONTH, levels = month.name),
+  mutate(
+    OCC_MONTH = factor(OCC_MONTH, levels = month.name),
     INJURY_COLLISIONS = as.factor(INJURY_COLLISIONS),
     FTR_COLLISIONS = as.factor(FTR_COLLISIONS),
-         PD_COLLISIONS = as.factor(PD_COLLISIONS),
-         AUTOMOBILE = as.factor(AUTOMOBILE),
-         PEDESTRIAN = as.factor(PEDESTRIAN))
+    PD_COLLISIONS = as.factor(PD_COLLISIONS)
+  )
 ```
@@
 # Introduction
-This paper analyzes traffic collision data in Toronto from 2014 to 2024.
-We explore trends in monthly, yearly, and neighborhood-level collisions, with a focus on variables such as fatalities, injuries, property damage, and modes of transport involved.
-The data is sourced from Open Data Toronto and contains detailed information about each collision.
+Traffic collisions remain a major public safety concern in urban areas, causing both human and economic costs.
+In cities like Toronto, understanding the temporal and categorical distribution of traffic collisions is crucial for reducing the incidence of accidents and improving road safety.
+Insights gained from such data allow policymakers and city planners to implement targeted interventions, especially during periods when collisions are most likely to occur.
 
-The remainder of this paper is structured as follows.
-Section @sec-data provides details of the dataset used in this analysis.
-Section @sec-results presents the results of the analysis, followed by a discussion in Section @sec-discussion.
-Finally, Section @sec-conclusion summarizes the key findings of the study.
+This paper examines traffic collisions in Toronto between 2014 and 2024.
+It aims to identify trends in monthly, daily, and hourly collision occurrences and assess the proportion of different collision types, including injuries, property damage, and failed to remain incidents.
+Fatalities are analyzed separately as a distinct trend.
+Analyzing these trends can guide the development of more effective road safety policies and public awareness campaigns.
+
+The key questions addressed in this paper are: (1) How do traffic collisions vary by month, day of week, and time of day?
+(2) What are the dominant collision types, and what do they reveal about driving behavior and road safety?
+(3) What are the long-term trends in fatalities, and how might they evolve in the future?
+
+This paper is organized as follows: @sec-data describes the data and methodology.
+@sec-results presents the results, including visualizations of key trends.
+Finally, @sec-discussion provides a detailed discussion of the findings and their implications for road safety and policy interventions.
 
 # Data {#sec-data}
-The dataset used in this analysis was obtained from Open Data Toronto.
-It includes traffic collision data collected between 2014 and 2024.
-The variables include the date and time of the collision, whether there were fatalities, injuries, or property damage, and the type of vehicle(s) involved.
-@tbl-1 provides a sample of the cleaned dataset.
+The data used in this paper came from the OpenData Toronto portal through the library `opendatatoronto` [@citeODT], specifically the Police Annual Statistical Report - Traffic Collisions [@CollisionDataCite].
+Data were cleaned and analyzed using the open source statistical programming language R [@citeR].
+Libraries `tidyverse` [@CiteTD], `knitr` [@knitr] and `dplyr` [@citedplyr] were used for simulating, cleaning and testing.
+Graphics were made using `ggplot2` [@citeGG].
+
+## Traffic Collision Analysis Data
+The cleaned dataset includes several key variables: Month, Day of the Week, Hour, Year, Fatalities, Injury Collisions, Failed to Remain (FTR) Collisions, and Property Damage (PD) Collisions.
+These variables allow for an analysis of collision frequency by time and type of incident.
+
+Certain variables such as Longitude, Latitude, Police Division, and Neighborhood information were removed as they were not necessary for this analysis.
+Additionally, vehicle type data was excluded since the majority of collisions involved automobiles, making this variable less useful for our purpose.
+
+@tbl-1 provides a random sample of the cleaned dataset, showing key variables such as the occurrence time and type of collision:
 
@@
 
-head(data, 10) |>
+set.seed(127)
+data %>%
+  sample_n(10) %>%
   kable(
-    col.names = c("Month", "Day of Week", "Year", "Hour", "Fatalities",
-                  "Injury Collisions", "FTR Collisions", "PD Collisions", 
-                  "Neighbourhood", "Automobile", "Motorcycle", 
-                  "Passenger", "Bicycle", "Pedestrian"),
-    booktabs = TRUE
+    col.names = c(
+      "Month", "Day of Week", "Year", "Hour", "Fatalities",
+      "Injury Collisions", "FTR Collisions", "PD Collisions"
+    ),
+    align = "c"
   )
@@
 # Results {#sec-results}
-Here we analyze the data by looking at the frequency of traffic collisions based on various factors such as month, year, and neighborhood.
 
-Collisions by Month
-The following figure shows the distribution of collisions by month across all years.
+## Collisions by Day of the Week
+@fig-collisions-day shows the distribution of traffic collisions by the day of the week.
+The data reveals that the highest number of collisions occurs on Friday, while Sunday has the lowest number of incidents.
+There is a significant difference between Friday and Sunday, with approximately 60% more collisions occurring on Fridays.
+During the week, Monday has the lowest number of collisions among weekdays, though it is still higher than weekends.
+Tuesday, Wednesday, and Thursday exhibit similar collision counts, which are slightly lower than Friday but higher than Monday.
+
+The overall trend suggests that traffic volume and collision risk peak towards the end of the workweek, likely due to higher road activity on Fridays as people prepare for the weekend.
+The decline on weekends, particularly on Sundays, can be attributed to lower traffic volume and fewer activities requiring vehicle use.
+
+```{r}
+#| label: fig-collisions-day
+#| fig-cap: Total traffic collisions by day of the week from 2014 to 2024
+#| echo: false
+
+data$OCC_DOW <- factor(data$OCC_DOW,
+  levels = c(
+    "Monday", "Tuesday", "Wednesday", "Thursday",
+    "Friday", "Saturday", "Sunday"
+  )
+)
+ggplot(data, aes(x = OCC_DOW)) +
+  geom_bar(position = "dodge", stat = "count", fill = "lightblue") +
+  theme_minimal() +
+  labs(
+    title = "Traffic Collisions by Day of the Week",
+    x = "Day of Week", y = "Total Count of Collisions"
+  ) +
+  theme(
+    axis.text.x = element_text(angle = 45, hjust = 1, color = "black"),
+    plot.title = element_text(hjust = 0.5, size = 14)
+  )
+```
+
+## Collisions by Month
+@fig-collisions-month shows the distribution of traffic collisions by month from 2014 to 2024.
+The data reveals a light W-shaped pattern, with collisions peaking at the start, middle, and end of the year.
+The highest counts occur in January and June, while the end of the year also sees an increase, though not as high as the other two peaks.
+This pattern could be linked to weather conditions, particularly during winter months like January when icy roads increase the likelihood of accidents.
+June's peak may coincide with higher traffic volumes during the summer as people travel for vacations and events.
+The increase toward the end of the year could reflect the impact of holiday travel and busy shopping periods.
 
@@
 #| label: fig-collisions-month
-#| fig-cap: Collisions by Month
+#| fig-cap: Monthly traffic collision counts in Toronto from 2014 to 2024.
 #| echo: false
 
-data %>% 
-  ggplot(aes(x = OCC_MONTH, fill = factor(OCC_YEAR))) +
-  geom_bar(position = "dodge") +
+ggplot(data, aes(x = OCC_MONTH)) +
+  geom_bar(position = "dodge", stat = "count", fill = "lightblue") +
   theme_minimal() +
-  labs(title = "Traffic Collisions by Month (2014-2024)",
-       x = "Month", y = "Count", fill = "Year") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  labs(
+    title = "Traffic Collisions by Month",
+    x = "Month", y = "Total Count of Collisions"
+  ) +
+  theme(
+    axis.text.x = element_text(angle = 45, hjust = 1, color = "black"),
+    plot.title = element_text(hjust = 0.5, size = 14)
+  )
 ```
 
-Collisions by Hour of the Day
+## Collisions by Hour of the Day
+@fig-collisions-hour shows the distribution of traffic collisions by hour of the day.
+The data indicates that collisions start to increase around 6 AM, coinciding with the start of the workday and upcoming rush hour traffic.
+The number of collisions continues to rise until it reaches a peak between 3 PM and 5 PM, during the afternoon rush hour, when roads are crowded with people returning home from work.
+After 6 PM, the number of collisions starts to decline as traffic decreases.
+The lowest collision rates are observed between midnight and 5 AM, when most people are asleep, and there are fewer vehicles on the road.
+This pattern is expected, as traffic volume and collision risk are closely correlated.
 
@@
 #| label: fig-collisions-hour
-#| fig-cap: Collisions by Hour of the Day
+#| fig-cap: Total collisions by hour of the day from 2014 to 2024
 #| echo: false
 
+data$OCC_HOUR[is.na(data$OCC_HOUR)] <- 0
 data %>%
-  ggplot(aes(x = OCC_HOUR, fill = factor(OCC_YEAR))) +
-  geom_histogram(binwidth = 1, position = "dodge") +
+  group_by(OCC_HOUR) %>%
+  summarise(Total_Collisions = sum(INJURY_COLLISIONS == "YES" | 
+                                     FTR_COLLISIONS == "YES" | 
+                                     PD_COLLISIONS == "YES", na.rm = TRUE)) %>%
+  ggplot(aes(x = OCC_HOUR, y = Total_Collisions)) +
+  geom_line(color = "steelblue", linewidth = 1) +
   theme_minimal() +
-  labs(title = "Traffic Collisions by Hour of the Day (2014-2024)",
-       x = "Hour", y = "Count", fill = "Year")
-
+  scale_x_continuous(breaks = 0:23) +
+  labs(
+    title = "Traffic Collisions by Hour of the Day",
+    x = "Hour of the Day",
+    y = "Total Number of Collisions"
+  ) +
+  theme(
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    plot.title = element_text(hjust = 0.5),
+    panel.grid.major.x = element_line(linewidth = 1.2),
+    panel.grid.major.y = element_line(linewidth = 0.5)
+  )
 ```
 
+## Collision Types
+@fig-collisions-type shows the proportion of each collision type.
+Property damage collisions dominate, accounting for approximately 71% of all recorded incidents.
+This suggests that most collisions involve vehicle-to-vehicle impacts or damage to roadside objects, without causing significant injuries.
+Failed to remain and injury collisions are relatively similar in proportion, each making up a smaller share of total collisions.
+Although these types can overlap (e.g., a failed-to-remain collision may also involve injury), the overlap is small enough to justify using a pie chart for this analysis.
+A more detailed analysis, using a Venn diagram, could account for these overlaps, but the overall trend remains clear: property damage is the most frequent outcome of traffic collisions.
+
 ```{r}
-#| label: fig-monthly-collisions
-#| fig-cap: Monthly distribution of traffic collisions
+#| label: fig-collisions-type
+#| fig-cap: Proportion of each type of traffic collisions (2014–2024).
 #| echo: false
 
-data %>%
-  count(OCC_MONTH) %>%
-  ggplot(aes(x = OCC_MONTH, y = n)) +
-  geom_bar(stat = "identity", fill = "skyblue") +
-  theme_minimal() +
-  labs(title = "Monthly Distribution of Traffic Collisions", 
-       x = "Month", y = "Number of Collisions")
+collision_summary <- data %>%
+  summarise(
+    Injury_Collisions = sum(INJURY_COLLISIONS == "YES"),
+    FTR_Collisions = sum(FTR_COLLISIONS == "YES"),
+    PD_Collisions = sum(PD_COLLISIONS == "YES")
+  )
 
+collision_summary_long <- collision_summary %>%
+  pivot_longer(
+    cols = everything(), names_to = "Collision_Type",
+    values_to = "Count"
+  ) %>%
+  mutate(Collision_Type = recode(Collision_Type,
+    "Injury_Collisions" = "Injury Collisions",
+    "FTR_Collisions" =
+      "Failed to Remain Collisions",
+    "PD_Collisions" =
+      "Property Damage Collisions"
+  ))
+
+collision_summary_long <- collision_summary_long %>%
+  mutate(Percentage = round(Count / sum(Count) * 100, 2))
+
+ggplot(collision_summary_long, aes(x = "", y = Count, fill = Collision_Type)) +
+  geom_bar(stat = "identity", width = 1) +
+  coord_polar("y", start = 0) +
+  geom_text(aes(label = paste0(Percentage, "%")),
+    position = position_stack(vjust = 0.5)
+  ) +
+  scale_fill_manual(values = c(
+    "Injury Collisions" = "lightblue",
+    "Failed to Remain Collisions" = "skyblue",
+    "Property Damage Collisions" = "#4682B4"
+  )) +
+  theme_void() +
+  labs(title = "Proportion of Each Collision Type") +
+  theme(
+    legend.title = element_blank(),
+    plot.title = element_text(hjust = 0.5)
+  )
 ```
 
+## Fatalities
+@fig-fatalities illustrates the trend in fatalities over time.
+The data shows fluctuations year by year, with the number of fatalities rising and falling between 2014 and 2024.
+However, the overall trend points to a gradual decline in fatalities, with the count for 2024 reaching its lowest point at 20 (as of late July).
+It is important to note that this number may rise as more data becomes available for the remainder of the year.
+The decline in fatalities may be attributed to improvements in road safety regulations, better vehicle technology, and more effective public safety policies.
+Nonetheless, the year-to-year fluctuations suggest that external factors, such as weather and traffic volume, continue to play a significant role in accident severity.
+
 ```{r}
-#| label: fig-fatalities-by-year
-#| fig-cap: Yearly trend in fatalities from collisions
+#| label: fig-fatalities
+#| fig-cap: Yearly trend in fatalities from collisions (2014-2024)
 #| echo: false
@@
   ggplot(aes(x = OCC_YEAR, y = Total_Fatalities)) +
-  geom_line(color = "darkred", size = 1) +
+  geom_line(color = "darkred", linewidth = 1) +
+  scale_x_continuous(breaks = seq(min(data$OCC_YEAR),
+    max(data$OCC_YEAR),
+    by = 1
+  )) +
   theme_minimal() +
-  labs(title = "Fatalities by Year", x = "Year", y = "Number of Fatalities")
+  labs(
+    title = "Fatalities by Year",
+    x = "Year",
+    y = "Number of Fatalities"
+  ) +
+  theme(
+    axis.text.x = element_text(angle = 45, hjust = 1),
+    plot.title = element_text(hjust = 0.5, size = 14),
+    panel.grid.major.x = element_line(linewidth = 0.8, color = "lightgrey")
+  )
+```
 
-```
 \newpage
 # Discussion {#sec-discussion}
-Overall, the analysis of traffic collisions from 2014 to 2024 reveals several key trends.
-Traffic collisions are more frequent during certain months of the year, with higher occurrences during the winter months, potentially due to hazardous road conditions.
-The distribution of collisions by hour of the day indicates that rush hours experience the highest number of collisions.
-The relationship between fatalities and injury collisions suggests that even non-fatal collisions result in significant injuries, emphasizing the need for road safety interventions.
+The analysis of traffic collisions in Toronto between 2014 and 2024 reveals several patterns, each offering insights into how urban life, driving habits, and external factors like weather affect road safety.
 
-# Conclusion {#sec-conclusion}
-This analysis highlights the importance of studying traffic collision data to inform policy decisions and road safety measures.
-Further analysis could include a more detailed breakdown by specific neighborhoods or examining the impact of weather conditions on collision frequency.
+## Day of the Week Analysis
+The distribution of traffic collisions across the days of the week provides a revealing look at how human behavior and traffic flow influence accident frequency.
+Friday, as the day with the highest collision rate, indicates a heightened risk toward the end of the workweek.
+This trend may be driven by multiple factors, including fatigue from the workweek, a possible rush to get home, and increased social activities as people transition into their weekend plans.
+Interestingly, Friday shows 60% more collisions than Sunday, a day characterized by reduced travel and leisure activities.
+The marked difference between these two days points to how daily schedules and behavioral changes significantly impact road safety.
+
+Sunday, with the lowest number of collisions, likely benefits from reduced road congestion, as many people tend to stay home or partake in local activities that don’t require long-distance travel.
+This also suggests that weekends, despite the occasional recreational travel, generally have safer road conditions.
+It is worth considering whether policies or public safety campaigns targeted at reducing Friday incidents could be beneficial, perhaps through encouraging more cautious driving during this high-risk period.
+
+## Monthly Patterns
+The light W-shaped pattern in monthly traffic collisions offers valuable insight into the seasonal and cultural influences that drive collision frequency.
+January, the peak month for collisions, aligns with Canada’s harsh winter conditions.
+The combination of icy roads, snowstorms, and reduced visibility presents a high risk for accidents.
+However, an additional contributing factor may be the increased road activity related to New Year’s celebrations, which often involve parties, alcohol consumption, and late-night travel.
+This brings attention to the need for targeted safety measures during major holidays, such as stricter policing of impaired driving.
+
+The second peak in June likely reflects the start of summer activities.
+Increased road travel for vacations, social events, and outdoor activities contributes to higher traffic volumes and, consequently, more collisions.
+The third rise at the end of the year, though not as pronounced, coincides with the busy holiday season when shopping, travel, and family visits dominate.
+Again, this suggests that city officials could benefit from strengthening public safety campaigns during these periods, with a particular focus on distracted driving, impaired driving, and road congestion management.
+As of last week, Toronto is addressing congestion with a plan that includes new fines, levies on construction projects, and automated enforcement, aimed at reducing traffic disruptions and improving road safety.
+These measures are crucial as the city's growing population and post-pandemic travel patterns have worsened congestion [@CongestionPlan].
+
+## Hourly Patterns
+Examining the distribution of collisions by hour of the day reveals a pronounced correlation between daily routines and road accidents.
+Collisions begin increasing sharply around 6 AM, as people wake up and head to work or school, contributing to morning rush hour congestion.
+This increase continues until 3 PM - 5 PM, when the evening rush hour dominates.
+The afternoon peak is particularly notable, reflecting the combined impact of people leaving work, running errands, and picking up children from school.
+The fact that this period consistently shows the highest number of collisions highlights the need for targeted road safety measures during rush hours, such as speed regulation, better traffic management systems, and public awareness campaigns focusing on attentive driving during high-congestion periods.
+
+After 6 PM, collisions decline drastically as road congestion eases and people settle into their evening routines.
+The midnight to 5 AM period records the fewest collisions, as expected, with minimal road activity during these hours.
+However, late-night incidents may still be influenced by riskier behaviors, such as driving under the influence of alcohol, making this period an important focus for law enforcement and awareness campaigns aimed at reducing late-night impaired driving.
+
+## Collision Types Implications
+The distribution of collision types adds depth to our understanding of how most accidents unfold.
+The dominance of property damage collisions—representing approximately 71% of all incidents—suggests that the majority of accidents are minor, involving either physical contact between vehicles or collisions with roadside infrastructure like signs or barriers.
+These accidents often result in fender benders or minor vehicle damage but don’t necessarily lead to injuries or fatalities.
+
+A smaller proportion of collisions involve injuries, where physical harm occurs to drivers or passengers.
+While these are slightly fewer in number compared to failed-to-remain collisions, they still represent a more severe subset of accidents that require further attention.
+Failed-to-remain collisions, where drivers flee the scene of the accident, are more common than injury collisions but have the least overlap with both injury and property damage incidents.
+This minimal overlap suggests that hit-and-run cases are a distinct category of accidents, often disconnected from the other two types, which might require different preventive strategies—such as better surveillance systems or harsher penalties for offenders.
+
+Given that fatalities are extreme outliers in terms of frequency, they warrant a separate analysis.
+While they occur in far fewer instances than injuries or property damage collisions, their severe consequences make them critically important for road safety planning.
+Fatal collisions often involve higher speeds, riskier driving behavior (such as driving under the influence), or particularly dangerous road conditions (e.g., icy roads).
+This reinforces the need for focused interventions targeting these outlier events.
+
+## Fatalities Insights
+The trend of declining fatalities from 2014 to 2024 is a positive outcome, reflecting the impact of improved road safety regulations, advances in vehicle technology (such as airbags, anti-lock brakes, and collision avoidance systems), and better public awareness of safe driving practices.
+However, it is important to acknowledge that this downward trend is influenced by a range of external factors, many of which are difficult to control.
+In a country like Canada, where winter conditions create significant challenges for drivers, weather-related accidents may remain an unfortunate reality.
+While safety campaigns can prepare drivers for icy or snowy conditions, accidents linked to poor weather may never be completely eliminated.
+
+Fatalities caused by controllable factors, such as distracted driving (e.g., using phones) or drunk driving, can be addressed more directly through stricter enforcement and educational campaigns.
+Policies that target these behaviors—such as harsher penalties for phone use while driving and regular sobriety checkpoints—can help reduce fatalities related to human error.
+While the 10-year period analyzed in this study reveals a favorable downward trend, it is a relatively short time frame for long-term safety assessments.
+Extrapolating from this data suggests that fatalities may continue to decline, but more robust, longitudinal data is needed to draw firmer conclusions.
+
+As of early September, the number of road fatalities in Toronto stands at 30 [@FatalitiesRecent], which still reflects a decline, but with a few months remaining in the year, it is unclear whether the final tally will continue this downward trend.
+This uncertainty underscores the importance of remaining vigilant and adaptable when it comes to road safety measures and policies, as the near future trend remains uncertain.
+Predicting future trends based on present data offers a useful tool for policymakers, but continual monitoring and adaptation of road safety strategies are essential to maintaining and enhancing this progress.
 

### 2024-09-27 paper/paper.qmd
update paper and citations

@@
   - Andy Jiang
-thanks: "Code and data are available at: https://github.com/AndyYanxunJiang/toronto-traffic-collisions.git"
-date: 26 September 2024
+thanks: "Code and data are available at: https://github.com/AndyYanxunJiang/toronto-traffic-collisions"
+date: today
 date-format: long
@@
 library(dplyr)
+library(here)
 
@@
 
-This paper examines traffic collisions in Toronto between 2014 and 2024.
-It aims to identify trends in monthly, daily, and hourly collision occurrences and assess the proportion of different collision types, including injuries, property damage, and failed to remain incidents.
+This paper examines traffic collisions in Toronto between 2014 and 2024, a decade that has seen substantial changes in road usage, vehicle technology, and city infrastructure.
+Despite the growing availability of traffic data, there is still a need for detailed analysis of collision patterns over time and by type.
+This paper seeks to fill this gap by identifying trends in monthly, daily, and hourly collision occurrences, and assessing the proportion of different collision types, including injuries, property damage, and failed to remain incidents.
 Fatalities are analyzed separately as a distinct trend.
-Analyzing these trends can guide the development of more effective road safety policies and public awareness campaigns.
+These insights can inform more effective road safety policies, particularly during high-risk periods.
 
@@
 # Data {#sec-data}
-The data used in this paper came from the OpenData Toronto portal through the library `opendatatoronto` [@citeODT], specifically the Police Annual Statistical Report - Traffic Collisions [@CollisionDataCite].
+The data used in this paper came from the Open Data Toronto portal through the library `opendatatoronto` [@citeODT], specifically the Police Annual Statistical Report - Traffic Collisions [@CollisionDataCite].
 Data were cleaned and analyzed using the open source statistical programming language R [@citeR].
-Libraries `tidyverse` [@CiteTD], `knitr` [@knitr] and `dplyr` [@citedplyr] were used for simulating, cleaning and testing.
+Libraries `tidyverse` [@citeTD], `knitr` [@citeknitr], `tibble` [@citetibble], `readr` [@citereadr], and `dplyr` [@citedplyr] were used for simulating, cleaning and testing.
+The package `here` [@citehere] was utilized for managing file paths to ensure reproducibility.
 Graphics were made using `ggplot2` [@citeGG].
+
+## Measurement
+The traffic collision data is collected by the Toronto Police and made publicly available through Open Data Toronto, where it is graded as Gold with a 100% data quality score.
+Each entry reflects a documented collision incident, capturing details like time, location, collision type, and vehicle type.
+However, the dataset does not include information on weather conditions or road quality, which could provide additional context for analyzing collision trends.
 
@@
 
-@tbl-1 provides a random sample of the cleaned dataset, showing key variables such as the occurrence time and type of collision:
+@tbl-1 provides a random sample of the cleaned dataset, displaying all relevant variables used for analysis:
 
@@
 #| label: tbl-1
-#| tbl-cap: Sample of the cleaned collision data
+#| tbl-cap: Random sample of the cleaned collision data
 #| echo: false
@@
 Failed to remain and injury collisions are relatively similar in proportion, each making up a smaller share of total collisions.
-Although these types can overlap (e.g., a failed-to-remain collision may also involve injury), the overlap is small enough to justify using a pie chart for this analysis.
+Although these types can overlap (e.g., a failed to remain collision may also involve injury), the overlap is small enough to justify using a pie chart for this analysis.
 A more detailed analysis, using a Venn diagram, could account for these overlaps, but the overall trend remains clear: property damage is the most frequent outcome of traffic collisions.
@@
 A smaller proportion of collisions involve injuries, where physical harm occurs to drivers or passengers.
-While these are slightly fewer in number compared to failed-to-remain collisions, they still represent a more severe subset of accidents that require further attention.
-Failed-to-remain collisions, where drivers flee the scene of the accident, are more common than injury collisions but have the least overlap with both injury and property damage incidents.
+While these are slightly fewer in number compared to failed to remain collisions, they still represent a more severe subset of accidents that require further attention.
+Failed to remain collisions, where drivers flee the scene of the accident, are more common than injury collisions but have the least overlap with both injury and property damage incidents.
 This minimal overlap suggests that hit-and-run cases are a distinct category of accidents, often disconnected from the other two types, which might require different preventive strategies—such as better surveillance systems or harsher penalties for offenders.

### 2024-09-27 scripts/00-simulate_data.R
include all library used

@@ -7,6 +7,8 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(tibble)
+library(readr)
 
 #### Simulate data ####
 set.seed(304)

### 2024-09-27 paper/example.qmd
usethis::git_vaccinate()

---
title: "More Civilian Deaths from Fires Linked to Lower Incomes in Toronto Wards"
author: "Carly Penrose"
thanks: "Code and data is available at: https://github.com/CarlyPenrose/Toronto_Fires."
date: today
date-format: long
abstract: "Hundreds of Canadians die and nearly one thousand are injured from fires each year. Some research suggests civilians in lower-income areas are more likely to die from fires than those in higher-income suburbs. Using Toronto Fire Service data from OpenData Toronto and Toronto ward profiles, this paper found that between 2018 and 2022, the most civilians died from fires in Toronto Centre. There appears to be a negative relationship between the median income of a ward and the number of civilian deaths from fires, with more people dying from fires in lower-income wards."
format: pdf
bibliography: ref.bib
---

```{r}
#| message: false
#| warning: false
#| echo: false

# set up workspace
library(tidyverse)
library(janitor)
library(knitr)
library(lubridate)
library(opendatatoronto)
library(scales)

# load the data
cleaned_fire_data <- read.csv("~/Toronto_Fires/outputs/Data/cleaned_fire_data.csv")
merged_ward_fires <- read.csv("~/Toronto_Fires/outputs/Data/merged_ward_fires.csv")
sum_per_ward <- read.csv("~/Toronto_Fires/outputs/Data/sum_per_ward.csv")
sum_per_year <- read.csv("~/Toronto_Fires/outputs/Data/sum_per_year.csv")
```

# Introduction

On December 24, 2023, a fire broke out in the West End of Toronto, [killing](https://www.cbc.ca/news/canada/toronto/rooming-house-fatal-fire-victim-1.7088988) William Cachia, a "well-loved" shelter worker [@cachiafire]. Earlier that same month, an unidentified man [died](https://www.cbc.ca/news/canada/toronto/forest-hill-fire-man-dead-1.7047400) in a house fire in a home that appeared to be abandoned in Toronto's Forest Hill neighbourhood [@foresthillfire]. Similarly tragic stories occur across the country, as hundreds of people are killed and nearly one thousand are injured in fires in Canada each year [@StatCanFires]. While structure fires can be devastating for all those who experience them, [news](https://www.thespec.com/news/hamilton-region/when-fire-meets-poverty-the-economics-of-fire/article_28c2acd7-2101-5579-b9cd-82b5368a09af.html) reports out of Hamilton, Ontario suggest fires are more deadly in low income areas compared to affluent suburban areas [@TheSpecCite]. A study in the United Kingdom also found that low-income people were more likely to experience deadly fires [@FireStudyUK].

The Toronto Fire Service (TFS) responds to thousands of fire events each year. The TFS is the largest fire service in Canada and the fifth-largest in [North America](https://www.toronto.ca/city-government/accountability-operations-customer-service/city-administration/staff-directory-divisions-and-customer-service/fire-services/) with 85 fire stations that serve a population of nearly 3 million people [@TFSCite]. The TFS keeps records of all of the incidents for which it is called and reports them to the Ontario Fire Marshall. The Ontario Fire Marshall provides this data to the city, which is then published on OpenData Toronto [@FireDataCite].

Toronto is separated into geographic areas known as wards. These electoral areas each have a population of roughly 100,000 people. The City of Toronto keeps records on demographic profiles of each of the city's 25 wards, including income metrics.

This paper tests the hypothesis that lower income areas have more dire outcomes from fires. First, it explores the number of fire incidents across years and across wards. Next, it looks at where and when civilian casualties from TFS-reported fire incidents were greatest from years 2018 to 2022. Finally, it plots the median household income in each ward according to 2021 census data against the number of civilians casualties from fires in that same ward to understand the relationship between median income and deaths from fires [@WardDataCite].

# Data

The data used in this paper came from the OpenData Toronto portal through the library `opendatatoronto` [@citeODT]. Data were cleaned and analyzed using the open source statistical programming language R [@citeR]. Libraries `tidyverse` [@CiteTD], `janitor` [@Citejan], `knitr` [@knitr], `dplyr` [@citedplyr], and `lubridate` [@Citelub] were used for simulating, cleaning and testing. Graphics were made using `ggplot2` [@citeGG].

## Fire Incidents data

Fire Incidents data are provided by the City of Toronto Fire Marshall and are updated on OpenData Toronto on an annual basis. However, the most recent data are from December 2022. The dataset tracks all fire incidents for which TFS was alerted, starting in 2011.

The dataset includes nearly 30,000 fire incidents over 11 years. It tracks the date of the incident, the location to the nearest intersection, the ward in which the incident took place, the time TFS was alerted to the fire and the time TFS arrived at the scene. It also records civilian casualties, TFS casualties, property lost (in dollars) and whether there was a working alarm on site.

In 2018, the City of Toronto switched from a 44-ward to a 25-ward model. Ward information recorded prior to 2018 would not match more recent data and could introduce errors into the analysis. Therefore, any incidents reported before Jan. 1, 2018 were excluded from this analysis.

The columns chosen for this analysis were: number of civilian casualties, TFS alarm time (used to identify the year of the incident) and incident ward. TFS alarm time was used to isolate the "year" variable because that data was more complete than the "incident time" data. Some potentially relevant columns were excluded. These include civilian injuries from fire incidents and the cost of damage from the fire. TFS firefighter casualties were not included in this analysis because, according to a 2014 [report](http://nfidcanada.ca/wp-content/uploads/2017/09/Fire-statistics-in-Canada-2005-to-2014.pdf) civilian casualties represent as many as 99% of deaths from structure fires, so the inclusion of firefighter deaths would be unlikely to change the analysis. [@NFIDstudy].

## Ward demographics

The City of Toronto provides ward profiles for each of its 25 wards, based on Statistics Canada data from the 2021 census [@WardDataCite]. Profiles include population, age, ethnic group, language spoken and median household income in a ward.  Median total household income is the amount of income a household makes from employment, investments and government supports and before taxes. The median income was used because it is less sensitive to extremely high and extremely low income values than the average.

Median income and civilian casualties by ward data were merged and plotted in a scatterplot to examine the relationship between the two variables.

# Results

## Fire incidents by year

There were, on average, nearly 3,500 fires incidents per year from 2018 to 2022. The most fire incidents occurred in 2022, followed closely by 2020 (@fig-1). The year with the fewest incidents was 2019, with just less than 3,000.

```{r}
#| label: fig-1
#| fig-cap: "Number of fire incidents in Toronto, by year, from 2018-2022"
#| warning: false
#| message: false
#| echo: false

cleaned_fire_data |>
  ggplot(aes(x = year)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Fire incidents in Toronto by year",
    x = "Year", y = "Fires"
  )
```

Despite the higher number of fires in 2022 and 2020, the greatest number of civilian deaths from fires happened in 2018 (@fig-2).

```{r}
#| label: fig-2
#| fig-cap: "Number of civilian casualties from fire incidents by year, 2018-2022"
#| warning: false
#| message: false
#| echo: false

ggplot(sum_per_year, aes(x = factor(year), y = deaths)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Civilian casualties from fires by year",
    x = "Year",
    y = "Civilian casualties"
  ) +
  theme_minimal()
```

## Fire incidents by ward

The greatest number of fire incidents happened in Ward 13, Toronto Centre. Ward 22, Scarborough Agincourt, had the fewest fires (@fig-3).

```{r}
#| label: fig-3
#| fig-cap: "Number of fire incidents in each ward, 2018-2022"
#| warning: false
#| message: false
#| echo: false

ggplot(cleaned_fire_data, aes(x = factor(ward))) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Fire incidents by ward",
    x = "Ward", y = "Fire incidents"
  )
```

Ward 13 saw the most civilians die in fires, while Ward 23, Scarborough North, had the fewest fire-related civilian casualties (@fig-4).

```{r}
#| label: fig-4
#| fig-cap: "Number of civilian casualties from fire incidents in each ward, 2018-2022"
#| warning: false
#| message: false
#| echo: false

ggplot(sum_per_ward, aes(x = factor(ward), y = total_casualties)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Civilian casualties from fires by ward",
    x = "Ward",
    y = "Civilian casualties"
  ) +
  theme_minimal()
```

## Fire incidents relationship to income

There appears to be a negative relationship between median income in a ward and the number of civilian casualties from fires in Toronto. The higher the median income of a ward, the fewer civilian deaths occur in that ward (@fig-5).

```{r}
#| label: fig-5
#| fig-cap: "Number of civilian casualties from fires in each ward by median income of the ward"
#| warning: false
#| message: false
#| echo: false

merged_ward_fires <- merged_ward_fires |>
  mutate(
    ward = as.factor(ward),
    median_income = as.numeric(median_income),
    total_casualties = as.integer(total_casualties)
  )

# build the scatter plot. Code taken from Telling Stories with Data Chapter 5
merged_ward_fires |>
  ggplot(mapping = aes(x = median_income, y = total_casualties, color = ward)) +
  geom_point() +
  scale_color_discrete() +
  # geom_smooth(method = lm, color = "black", se = FALSE) +
  # code to show labels on the plot points referenced from https://github.com/InessaDeAngelis/Toronto_Elections/blob/main/outputs/paper/toronto_elections_paper.qmd
  geom_text(aes(label = ward), vjust = -0.6, hjust = 0.75, size = 2) +
  labs(
    title = "Civilian casualties from fires by ward income",
    x = "Median ward income (2020)",
    y = "Civilian casualties",
    color = "Ward"
  ) +
  theme_minimal() +
  # this section was taken from ChatGPT to ensure the x-axis values were shown in full vs mathematical notation. I tried code from various places, including http://www.sthda.com/english/wiki/ggplot2-axis-scales-and-transformations#format-axis-tick-mark-labels-1 and from previous student work, but wasn't able to come up with a way to display the income numbers properly without LLM help.
  scale_x_continuous(
    breaks = seq(min(merged_ward_fires$median_income, na.rm = TRUE),
      max(merged_ward_fires$median_income, na.rm = TRUE),
      by = 5000
    ),
    labels = scales::number_format(scale = 1e-3, suffix = "k")
  )
```

In Toronto Centre, which has the lowest median income in the city at $65,000, 52 people died from fires between 2018 and 2022. The ward with the highest median household income was Ward 25, Scarborough-Rouge Park (median income $105,000), which includes the high-income Bridle Path neighbourhood. In that ward, only 11 civilians died in fires.

The ward where the fewest people died from fires was Scarborough North which has a median household income of $87,000. Only eight people died from fires over four years.

# Discussion

This paper investigated the wards where the most civilians died from fire incidents, according to TFS data retrieved from OpenData Toronto. Results show that more civilians died from fires in Ward 13, Toronto Centre, than any other ward over the years of 2018 to 2022. The fewest civilians died from fires in Ward 23, Scarborough North. There appeared to be a negative relationship between number of civilian casualties from fires and the median income of a ward, with a lower median income correlating to more civilian casualties from fires.

It is possible this relationship could be explained by other factors such as density of housing units or traffic. There is also research to suggest that more affordable housing may be more likely to lack updated fire safety standards for tenants [@firesafety]. More research should be done to clarify the nature of the relationship between income and civilian risk of death in a fire.

It is also possible other factors, like TFS response time, are correlated to more deaths. The dataset used for this analysis also provides the time TFS was alerted to an incident as well as the time they arrived on site. Examining how the length of time to respond is related to casualties and the relationship to median income could further this research. Other demographic factors are linked to lower-income due to systemic bias and discrimination like ethic background or immigration status. These factors could be another dimension to study these data.

These findings could help inform policy outside of just fire prevention education. If further research supports this finding, it could inform public safety steps local governments could take. As one example, local governments could conduct more routine and widespread fire safety inspections in low-income areas. It could also concentrate more firefighters in these problem areas. While prevention is key, understanding how to improve the response to fires could help save lives.

## Limitations 
Importantly, the TFS data should be verified for accuracy. TFS fire incident data used, 173 people died from fires in 2018. However, according to some Statistics Canada reports, only 161 people died from fires across the country. It would be important to compare the data collection methods for the TFS to the methods used by Statistics Canada to ensure the data are compiled consistently and correctly.

The TFS data were incomplete. Data for 144 incidents had to be excluded from the analysis because they did not include the ward in which the incident took place. It is unlikely, however, that these data would change the outcome of the analysis, as according to the Ontario Fire Marshall, incidents with incomplete data are either under investigation or classified as a "no loss outdoor fire," which would have resulted in no civilian casualties.

Additionally, for the scope and purposes of this analysis, the only variable used to gauge the severity of fires was civilian deaths. But TFS data includes other variables including injuries and costs of fire-related damage to property. An analysis that includes these variables could be an important additional measure of the severity of fires.

# References

### 2024-09-27 paper/paper.qmd
polish paper

@@
 ---
-title: "Traffic Collisions in Toronto: Analysis of Key Variables"
-subtitle: "An Analysis of Traffic Collision Data from 2014-2024"
+title: "Traffic Collisions in Toronto"
+subtitle: "Analysis of Trends and Safety Insights"
 author: 
   - Andy Jiang
-thanks: "Code and data are available at: https://github.com/AndyYanxunJiang/toronto-traffic-collisions"
+thanks: "Code and data are available at: https://github.com/AndyYanxunJiang/toronto-traffic-collisions."
 date: today
@@
 
-@tbl-1 provides a random sample of the cleaned dataset, displaying all relevant variables used for analysis:
+@tbl-sample provides a random sample of the cleaned dataset, displaying all relevant variables used for analysis:
 
 ```{r}
-#| label: tbl-1
-#| tbl-cap: Random sample of the cleaned collision data
+#| label: tbl-sample
+#| tbl-cap: Random sample of 10 rows from the cleaned collision data
 #| echo: false
@@
 #| label: fig-collisions-day
-#| fig-cap: Total traffic collisions by day of the week from 2014 to 2024
+#| fig-cap: Total traffic collisions by day of the week in Toronto from 2014 to 2024.
 #| echo: false
@@
 #| label: fig-collisions-month
-#| fig-cap: Monthly traffic collision counts in Toronto from 2014 to 2024.
+#| fig-cap: Total traffic collisions by month in Toronto from 2014 to 2024.
 #| echo: false
@@
 #| label: fig-collisions-hour
-#| fig-cap: Total collisions by hour of the day from 2014 to 2024
+#| fig-cap: Total traffic collisions by hour of the day in Toronto from 2014 to 2024.
 #| echo: false
@@
 #| label: fig-collisions-type
-#| fig-cap: Proportion of each type of traffic collisions (2014–2024).
+#| fig-cap: Proportion of each type of traffic collisions in Toronto from 2014 to 2024.
 #| echo: false
@@
 #| label: fig-fatalities
-#| fig-cap: Yearly trend in fatalities from collisions (2014-2024)
+#| fig-cap: Yearly trend in fatalities from traffic collisions in Toronto from 2014 to 2024.
 #| echo: false
@@
 
-Sunday, with the lowest number of collisions, likely benefits from reduced road congestion, as many people tend to stay home or partake in local activities that don’t require long-distance travel.
+Sunday, with the lowest number of collisions, likely benefits from reduced road congestion, as many people tend to stay home or partake in local activities that don't require long-distance travel.
 This also suggests that weekends, despite the occasional recreational travel, generally have safer road conditions.
@@
 The dominance of property damage collisions—representing approximately 71% of all incidents—suggests that the majority of accidents are minor, involving either physical contact between vehicles or collisions with roadside infrastructure like signs or barriers.
-These accidents often result in fender benders or minor vehicle damage but don’t necessarily lead to injuries or fatalities.
+These accidents often result in fender benders or minor vehicle damage but don't necessarily lead to injuries or fatalities.
 

### 2024-09-27 paper/paper.qmd
add table of content

@@
 format: pdf
+toc: true
+number-sections: true
 bibliography: references.bib
